---
title: "TITLE"
categories:
  - tidytuesday
date: "2026-03-24"
draft: true
execute:
  echo: true
  message: false
  warning: false
format: 
  html:
    code-fold: true
---

Link to data: [TidyTuesday](https://github.com/rfordatascience/tidytuesday/blob/main/data/2026/2026-08-25/readme.md)

> Today's video is a DEEP DIVE into the world of country music, and I'm determined to figure out if the stereotypes people have about modern country songwriting are actually true. Is it REALLY all about beer and trucks and girls and blue jeans? Are all the lyrics REALLY the same? To find out, I maintained a spreadsheet of every country song that has reached the Top 30 of Billboard's Country Airplay chart for the past six years (2014-2019) and then crunched a whole lot of data.


## 0. Data

```{r}
#| code-fold: false
# Packages
library(tidyverse)

# Load data
tuesdata <- tidytuesdayR::tt_load('2026-08-25')

country_lyrics <- tuesdata$country_lyrics
top_all_writers <- tuesdata$top_all_writers
top_primary_writers <- tuesdata$top_primary_writers
top_producers <- tuesdata$top_producers
```

# Visualization 1

## 1. Data Cleaning

```{r}
library(tidytext)
library(stopwords)
# Data set of weighted word proportions
word_props <- country_lyrics |> 
  group_by(song, artist) |> 
  unnest_tokens(word, lyrics) |> # Gets word by song and artist 
  filter(!(word %in% stop_words$word)) |> # Remove stop words
  group_by(song, artist, word) |> 
  count(sort = T) |> # Counts of words within songs
  group_by(artist) |> 
  complete(song, word, fill = list(n = 0)) |> # For each artist, fill in 0's for words not in their other songs
  
  group_by(song, artist) |> 
  mutate(prop_in_song = n/sum(n)) |>  # Get the proportion within a song
  group_by(artist, word) |> 
  summarise(weighted_prop = mean(prop_in_song)) |> # Get the proportion across songs
  arrange(-weighted_prop) 

# Data set of songs by artist
artist_counts <- country_lyrics %>% 
  group_by(artist) %>% 
  count(sort = T) %>% 
  filter(n >= 10) # Filter artists with at least 10 sings in the top hits

# Inner join to filter
country <- word_props %>% 
  inner_join(artist_counts, by = 'artist')

```


## 2. Data Visualization

```{r}
library(ggwordcloud)
country %>% 
  ungroup() %>% 
  filter(weighted_prop > 0.001) %>% 
  ggplot(mapping = aes(label = word, 
                       size = weighted_prop,
                       color = artist)) + 
  geom_text_wordcloud(area_corr = TRUE) + 
  scale_size_area(max_size=24, trans = power_trans())
```


```{r}

```


# Visualization 2

## 1. Data Cleaning

```{r}

```


## 2. Data Visualization

```{r}

```