---
title: "Nova Scotia Ocean Temperatures"
categories:
  - tidytuesday
date: "2026-03-31"
draft: true
execute:
  echo: true
  message: false
  warning: false
format: 
  html:
    code-fold: true
---


This #tidytuesday comes from March 31st, 2026 and explores ocean temperature in Nova Scotia. The GitHub page can be found [here](https://github.com/rfordatascience/tidytuesday/tree/main/data/2026/2026-03-31) and provides the following description: 

> The Province of Nova Scotia recognizes the importance of coastal waters, which are critical to the prosperity and sustainability of rural and coastal communities. To bridge the gap between science and decision making, the Nova Scotia Department of Fisheries and Aquaculture (NSDFA) partners with the Centre for Marine Applied Research (CMAR) to measure the environmental conditions of Nova Scotia’s coastal waters.

My goal with these datasets is as follows: 

1. Create a media-quality graphic 
2. Create a statistical graphic


## Data

```{r}
# Packages
library(tidyverse)

# Load data
tuesdata <- tidytuesdayR::tt_load('2026-03-31')

# Extract data
ocean_temperature <- tuesdata$ocean_temperature
ocean_temperature_deployments <- tuesdata$ocean_temperature_deployments
```

## Media Graphic

To show where ocean temperature sensors are positioned around Nova Scotia, I decided to use a map-based graphic. Plotting deployment locations over the Canada map gives immediate geographic context for where these measurements were collected.

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-dpi: 1200

library(leaflet)

canada <- map_data('world', 'canada')

canada %>% 
  # filter(subregion %in% c('Cape Breton Island', 'Sable Island')) %>% 
  ggplot(mapping = aes(x = long, y = lat, group = group)) + 
  geom_polygon() +
  geom_point(aes(x = longitude, y = latitude, group = NULL), data = ocean_temperature_deployments,
             color = 'red') + 
  theme_void() +
  coord_map()

```

## Statistical Graphic

The ocean temperature dataset includes repeated measurements over time at multiple sensor depths. This chart shows how mean temperature changes over time across those depth levels.



```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-dpi: 1200

ocean_temperature %>% 
  ggplot(mapping = aes(x = date, y = mean_temperature_degree_c)) + 
  geom_line() + 
  facet_wrap(~sensor_depth_at_low_tide_m)



```

