---
title: "One Million Digits of Pi"
categories:
  - tidytuesday
date: "2026-03-24"
execute:
  echo: true
  message: false
  warning: false
format: 
  html:
    code-fold: true
---

Pi is an irrational number representing the ratio of a circle's circumference to its diameter. For this week's [TidyTuesday](https://github.com/rfordatascience/tidytuesday/tree/main/data/2026/2026-03-24), one million digits of pi and their placements are provided for a fun exploration project. 

> "Pi is an irrational number, meaning its decimal representation never ends and never settles into a permanent repeating pattern."" — Eve Andersson, collector of the one million digits dataset

I saw a great post by [Steven Ponce on Bluesky](https://bsky.app/profile/sponce1.bsky.social/post/3mhlvttmhis23) showing the walking path of pi using the the digits to as compass directions. I decided to recreate the concept of this visualization, as well make my own contribution.

# Walking Path of π

## 1. Reading in the data

```{r}
# Packages
library(tidyverse)

# Load data
tuesdata <- tidytuesdayR::tt_load('2026-03-24')

# Extract data
pi_digits <- tuesdata$pi_digits

# Show the first few values
head(pi_digits) %>% 
  knitr::kable()
```

## 2. Data Cleaning

A compass has 360 degrees, which can be divided into 10 increments of 36 degrees. Assigning digits to these increments provide directions, where 0 = 0°, 1 = 36°, ..., and 9 = 324°. 

With directions assigned, the next step is to figure out the positioning of each step. If $\theta$ represents the direction, the X-Y displacement is given by:

- x-displacement: $\cos(\theta)$
- y-displacement: $\sin(\theta)$

Here, R requires that $\theta$ be represented by radians, meaning that we first need to transform our degrees, $D$ into radians using $D\times \frac{\pi}{180}=\theta$. After getting the X and Y displacements, the cumulative sum along the digits 

```{r}
# Define the direction in degrees and radians
step_directions <- data.frame(
  digit = 0:9,
  degree = seq(0, 360-36, l = 10) #Note: 360 is the same as 0
) %>% 
  mutate(
    radian = degree*pi/180
  )

# Set end point to set the number of digits
end_spot <- 10000

# Create the walking path
pi_walk <- pi_digits %>% 
  left_join(step_directions, by = 'digit') %>% 
  mutate(displacement_x = cos(radian),
         displacement_y = sin(radian),
         pos_x = cumsum(displacement_x),
         pos_y = cumsum(displacement_y)) %>% 
  filter(digit_position <= end_spot)
```


## 3. Visualization

```{r}
p_walk <- pi_walk %>% 
  ggplot(mapping = aes(x = pos_x, y = pos_y)) + 
  geom_path(alpha = 1/3, color = '#63B3ED') + 
  
  # Aethestics
  labs(title = "Where did my π go?",
       subtitle = 'The walking path of π. Digit values are used to\nmark the direction of travel (0=0°, 1=36°, ..., 9=324°)',
       caption = '') + 
  theme_void() + 
  theme(aspect.ratio = 1,
        plot.background =  element_rect(fill = '#151D28', color = 'black'),
        plot.margin = margin(1/2, 1, 0, 1/2, "cm"),
        plot.title = element_text(color = 'grey99', face = 'bold', size = 24),
        plot.subtitle = element_text(color = 'grey99', size = 10))
p_walk
```

## 4. Adding Animations

The static chart above shows the complete path, but we can make it more engaging by animating it with the `gganimate` package. I also included a point to clearly show the current position of the walking path:

```{r}
# Load package
library(gganimate)

# Update previous plot
p_animate <- p_walk + 
  geom_point(color = 'gold', size = 2) + 
  transition_reveal(digit_position) +
  ease_aes('linear')

# Animate!
animate(p_animate, nframes = 100, end_pause = 10, rewind = TRUE)
```

# Bump Chart

Many visualization of pi tend to focus on the distributions of digits. While this gives an idea of how often a specific digit is represented, we can show how this changes over time with a bump chart for the ranking of counts. 

## 1. Data Cleaning

```{r}
#| cache: true

# Set upper limit of digits (for testing purposes of smaller dataset)
n_digits <- nrow(pi_digits)

# Get all digits for all positions (10 million rows!)
pi_markers <- pi_digits %>% 
  filter(digit_position <= n_digits) %>% 
  expand(digit_position, digit2 = digit)

# Get cumulative counts at each position and rank by count
pi_ranks <- pi_digits %>% 
  filter(digit_position <= n_digits) %>% 
  full_join(pi_markers) %>% 
  mutate(match = as.numeric(digit == digit2)) %>% 
  group_by(digit2) %>% 
  mutate(total = cumsum(match)) %>% 
  select(digit_position, digit2, total) %>% 
  group_by(digit_position) %>% 
  mutate(rank = rank(-total, ties.method = 'first')) #negative so 1 = largest count

# Remove markers to clear some memory
rm(pi_markers)
```

## 2. Visualization

The `ggbump` package works well here, but is no longer available on CRAN. Either the archive version of this package can be downloaded, or the development version from [GitHub](https://github.com/davidsjoberg/ggbump).

A clear issue with 1 million positions is that the there will be 10 million data points. This is excessive, so I decided to only show ranks along the log base 10 scale. This seems to add clarity in the resulting chart, which is better than showing every value. 

```{r}
#| fig-dpi: 300
#| fig-width: 6

#devtools::install_github("davidsjoberg/ggbump")
library(ggbump)
library(scales)

breaks <- 10*10^c(0:10)

p_bump <- pi_ranks %>% 
  filter(digit_position %in% breaks) %>%
  
  # Base chart
  ggplot(mapping = aes(x = digit_position, y = rank, 
                       group = digit2, color = factor(digit2))) + 
  geom_bump() + 
  geom_point(size = 3) + 
  geom_text(aes(label = digit2), color = 'black', size = 2) + 
  
  # Aesthetics
  scale_y_reverse(breaks = 1:10) + 
  scale_x_log10(breaks = breaks, labels = comma) + 
  labs(
    x = 'Digit Position',
    y = 'Rank',
    title = 'Digit Frequency Rankings For Pi',
    subtitle = 'Ranks are computed as the cumulative count of digits\nat equally spaced intervals of the log 10 scale.'
  ) + 
  theme_minimal() + 
  theme(
    aspect.ratio = 1/2,
    legend.position = 'none',
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(color = 'grey70'),
    plot.title = element_text(face = 'bold', size = 16),
    plot.subtitle = element_text(size = 10, color = 'grey20'),
    axis.title = element_text(size = 12, color = 'black'),
    plot.background = element_rect(color = 'black', fill = 'grey90')
  )

p_bump
```




