---
title: "TITLE"
categories:
  - tidytuesday
date: "2026-03-24"
draft: true
execute:
  echo: true
  message: false
  warning: false
format: 
  html:
    code-fold: true
---

Link to data: [TidyTuesday](https://github.com/rfordatascience/tidytuesday/blob/main/data/2026/2026-08-11/readme.md)

> We use the sample of emission-line nuclei derived from a recently completed optical spectroscopic survey of nearby galaxies to quantify the incidence of local (z = 0) nuclear activity. [...] Half of the objects can be classified as H II or star-forming nuclei and the other half as some form of AGN, of which we distinguish three classes — Seyfert nuclei, LINERs, and transition objects.

## 0. Data

```{r}
#| code-fold: false
# Packages
library(tidyverse)

# Load data
tuesdata <- tidytuesdayR::tt_load('2026-08-11')

palomar_emission_lines <- tuesdata$palomar_emission_lines
palomar_survey <- tuesdata$palomar_survey
```

# Visualization 1

## 1. Data Cleaning

```{r}
joined_data <- palomar_survey |> 
  full_join(palomar_emission_lines, by = 'galaxy_name') |> 
  filter(activity_type != 'Absorption' & !is.na(activity_type))
names(joined_data)
```


## 2. Data Visualization

```{r}
#| fig-width: 6
#| fig-height: 4
# joined_data |>
palomar_survey |>
  filter(activity_type != 'Absorption' & !is.na(activity_type)) |> 
  ggplot(mapping = aes(x = log_nii_ha, y = log_oiii_hb, shape = activity_type, color = activity_type)) + 
  geom_point() + 
  scale_x_log10() +
  scale_y_log10() +
  theme_bw() + 
  theme(aspect.ratio = 4/6)

```

